---
title: Discrimination among groups -- discriminant analysis (DA), classification tree
  analysis (CART) and indicator species analysis (ISA)
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(vegan)
```

This info is taken from Andreas Hamann's class at the University of Alberta, [Multivariate Statistics for the Environmental Sciences](https://sites.ualberta.ca/~ahamann/teaching/renr690/).

# Discriminant analysis

Discriminant analysis includes metric-based techniques (as opposed to non-metric). You can think of a Canonical discriminant analysis (CANDISC) as the metric equivalent of a NMDS for a non-metric application: that is, it reduces complexity amongst axes of variation. Similarly, you can use a Multivariate analysis of variance (MANOVA) to determine the effects of different predictor variables, discrimination analysis to classify groups (i.e. answer the question:w hat differentiates classes?) and DISCRIMc to make predictions (i.e. answer the question, what group does this new species fit into).

## Canonical discriminant analysis (CANDISC)

CANDISC is very similar to PCA in that the coordinate system is rotated to maximize variance,m but now for CANDISC the variance amongst _groups_ is maximized, while for PCA the variance amongst _observations_ is maximized. This is a _canonical_ procedure, which means that once the first axis is fixed, the next thing is to fix the next axis, then the next, etc.  

If the objective is to reduce complexity (as you would want with a PCA) but also show groups and demonstrate what variables contribute to differences between those groups, then CANDISC is for you.  

For discriminant analysis, the maximum number of discriminant functions (i.e. the number of axes) is much less than what you would get from a PCA - for example, if you have 17 different variables in your PCA, you could have 17 different principal components; however, the maximum number of axes in CANDISC is the number of groups minus 1. This is because regardless of how many variables your data has, the groups are really only explained by the potential number of groups. If you only have 2 groups, you can't actually do a CANDISC.  

If you want to predict group membership, you can do that with a CANDISC. If you have a new species, you can find its coordinates in the CANDISC, and then measure the Mahalanobis distance (i.e. the smalles normalized Euclidean distance to the centroid) for that point, and the  smallest distance indicates the group to which that species belongs.  


















